Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-05-23 paper/meg_exps.tex
supplementary

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_sensors.pdf}
  \caption{Magnetosensor response (femtoteslas fT) averaged across words for the first subject. The color coding corresponds to the positions of the sensors on the head as shown on the top-left diagram. The diverging curves correspond to artifacts in the MEG data.}
  \label{fig:megavg}
\end{figure}

@@

We attempt to identify which factors of $X$ explain the MEG
recordings $Y$ at each time sample and for each subject separately. To this aim,
we compare three methods: forward regression, backward regression, and back-to-back regression.
Corresponding results can be found in Figure \ref{fig:meg_twocurves}.

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
  \caption{We applied the forward regression (a.k.a encoding coefficients), the backward regression (a.k.a decoding scores) and B2B on real MEG data. Top row: the first subject. Bottom row: average across subjects.
  We plot the contributions of the \textit{word frequency} and the \textit{word
  length} over time. Our method can disambiguate them consistently with
  previous neuroscientific work showing an early response ($\approx$ 150ms) to
  length and a later response ($\approx$ 400ms) to frequency.}
  \label{fig:meg_twocurves}
\end{figure}

Forward regression performs a regularized least-squares regression $A \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| A_{i, :} \|^2$.
%
Backward regression performs a regularized least-squares regression $B \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YB$, using a validation sample.
%
Finally, we used the back-to-back regression as described earlier, using a bagging of 10 alphas logarithmically distributed between -5 and 5.

@@

The three types of analyses were implemented for each subject and each time
sample independently. To estimate the reliability of these estimates in the
population, we tested, using one-sample $t$-test whether the distribution of
the estimated contribution of each factor different from baseline ($t=0$ms),
as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ ranging from
150ms to 1s after word onset. We use the resulting p-values as proxy to quantify
the performance of each model. These results are depicted in Figure
\ref{fig:meg_pval}. For concision purposes, we only report this metric averaged
between 150ms and 1s after word onset.

@@

In Figure \ref{fig:meg_twocurves} we show, for each model, the effects obtained
in a representative subject, as well as the average effects across subjects.
The results show that all models significantly detect that both word length and
word frequency modulate brain responses after the stimulus onset. However
forward regression detected word length ($t-value=10.6$, $p=10^{-16}$) and
word frequency ($t-value=11.1$, $p=10^-17$) less reliably than B2B (length:
$t-value=18.7$, $p=10^{-31}$, frequency:$t-value=15.9$, $p=10^{-26}$). B2B
was less sensitive than the backward regression (length: $t-value=33.6$,
$p=10^{-50}$, $t-value=31.7$, frequency:$p=10^{-48}$). However, as detailed in
the introduction, backward models confound collinear factors. Consequently,
the high sensitivity of our backward regression is likely to come from a loss of specificity. This possibility is corroborated by the fact that the dynamics
associated with word length and word frequency are remarkably similar in the
backward model.

>>>>

@@

We attempt to identify which word features ($X$) is reliably associated with the MEG recordings $Y$ at each time sample and for each subject separately. To this aim,
we compare three methods: forward regression, backward regression, and back-to-back regression.
Corresponding results can be found in Figure \ref{fig:meg_twocurves}.

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
  \caption{We applied the forward regression (a.k.a encoding coefficients), the backward regression (a.k.a decoding scores) and B2B on real MEG data. Top row: the first subject. Bottom row: average across subjects.
  We plot the contributions of the \textit{word frequency} and the \textit{word
  length} over time. B2B can disambiguate these two factors consistently with
  previous neuroscientific work \cite{kutas2011thirty}\citep{pegado2014timing}
  showing an early response ($\approx$ 150ms) to length and a later response
  ($\approx$ 400ms) to frequency.}
  \label{fig:meg_twocurves}
\end{figure}

Forward regression was performed with an l2-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
%
Backward regression was performed with an l2-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
%
Finally, we used the back-to-back regression was performed using two l2-regularized
least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging of 10 alphas logarithmically distributed between -5 and 5.

@@

The three types of analyses were implemented for each subject and each time
sample independently. To estimate the reliability of these estimates in the
population, we tested, using one-sample $t$-test whether the distribution of
the estimated contribution of each factor different from baseline ($t=0$ms),
as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ ranging from
150ms to 1s after word onset. We use the resulting t and p-values as proxy to
quantify the performance of each model. These results are depicted in Figure
\ref{fig:meg_pval}. For concision purposes, we only report this metric averaged
between 150ms and 1s after word onset.

@@

In Figure \ref{fig:meg_twocurves} we show, for each model, the effects obtained
in a representative subject, as well as the average effects across subjects.
The results show that all models significantly detect that both word length and
word frequency modulate brain responses after the stimulus onset. However
forward regression detected word length ($t$-value$=10.6$, $p=10^{-16}$) and
word frequency ($t$-value$=11.1$, $p=10^-17$) less reliably than B2B (length:
$t$-value$=18.7$, $p=10^{-31}$, frequency:$t$-value$=15.9$, $p=10^{-26}$). B2B
was less sensitive than the backward regression (length: $t$-value$=33.6$,
$p=10^{-50}$, $t$-value$=31.7$, frequency:$p=10^{-48}$). However, as detailed in
the introduction, backward models confound collinear factors. Consequently,
the high sensitivity of our backward regression is likely to come from a loss of specificity. This possibility is corroborated by the fact that the dynamics
associated with word length and word frequency are remarkably similar in the
backward model.
